---
title: "Covid19_Data_Review"
author: "CU Learner 2023"
date: "`r Sys.Date()`"
output: html_document
---

## Project #2: Covid Data Review

This report captures the process followed to import, tidy and prepare covid time series data.  I will be using this data for high level visualization, analysis and modeling.  This data is provided by Johns Hopkins and can be accessed at the following github repository:
<https://www.github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/>

***Note: the raw files were utilized for import of data***

The objective of this report is to better understand:

1. The trend of cases and deaths in the United states over time.
2. What areas experienced the highest and lowest incidence of recorded infection and death.
3. What a linear model would indicate for outcomes in the United States, specifically deaths as a function of number of cases in each state and relative to population size.
4. What areas managed outcomes the best and worst when compared with the linear model.


```{r setup, include=FALSE}
# Loading Essential Libraries
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)

```

The libraries used in this report are listed below:

* library(tidyverse) 
* library(lubridate)

### Importing the Data

The first step will be to import the data from the Johns Hopkins github repository.
I have followed the same process as shown in the lectures.

```{r jh_data_import, echo = TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_recovered_global.csv")

urls <- str_c(url_in, file_names)

#Source data urls shown below
urls

#reading data from csv into local variables
us_cases <- read_csv(urls[1], show_col_types = FALSE)
global_cases <- read_csv(urls[2], show_col_types = FALSE)
us_deaths <- read_csv(urls[3], show_col_types = FALSE)
global_deaths <- read_csv(urls[4], show_col_types = FALSE)
global_recovered <- read_csv(urls[5], show_col_types = FALSE)

```
### Tidying the Data
See comments to understand what has been done in each code chunk to "wrangle" the data prior to analysis.
```{r tidy_data, echo = TRUE}
#Removing lat / long columns because they will not be used in this study
#Pivoting dates from columns to rows

global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))

global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long))

#Merging cases and deaths into single variable
#Renaming headers for Province/State and Country/Region
#Converting from character to date format for date values
global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>% mutate(date = mdy(date)) 

#Filter global data to show only records having at least one case
global <- global %>% filter(cases > 0)

# Performing same transformation to US cases and deaths
# Pivoting date columns to rows and changing values to date format
# Eliminating columns that will not be used for analysis
us_cases <- us_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))

us_deaths <- us_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))

# Merging data for US cases and deaths
US <- us_cases %>% full_join(us_deaths)

#Filter US data to show only records having at least one case
US <- US %>% filter(cases > 0)

global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url, show_col_types = FALSE) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

```
### Visualizing Data

The datasets are first being grouped in order to aid in visualizing, analyzing and modeling the data.

```{r group_data, echo = TRUE}
# Consolidating cases and deaths by state and adding column for deaths per million
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

# Consolidating cases and deaths for the US by date, and adding a column for deaths per million
US_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
```

The US data is now grouped at the state and national level.

Below is a visualization of cases and deaths at the national level over time.

```{r visual_data, echo = TRUE, messages = FALSE, warning = FALSE}
US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y = NULL)

```

The code chunk below will make use of two new variables enabling visualization of new cases and new deaths over time.

```{r visual_data_new, echo = TRUE, messages = FALSE, warning = FALSE}
# Adding columns with value of new cases and deaths for state and national data sets
US_by_state <- US_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

US_totals %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y = NULL)

global <- global %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

```

Below, I create the same filter to isolate a single state as shown in the lecture.

```{r state_filter, echo = TRUE}

state <- "New York"
US_select_state <- US_by_state %>% filter(Province_State == state, new_cases > 0, new_deaths > 0)

US_select_state %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID19 in ",state), y = NULL)

```

The following code chunk is grouping the data by Province_State to facilitate easy comparison of deaths and cases per thousand people.  

```{r grouping_states, echo=TRUE}
US_state_totals <- US_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases / population, deaths_per_thou = 1000 * deaths / population) %>% filter(cases > 0, population > 0)
```

The tables that follow show what areas experienced the highest and lowest incidence of cases and deaths.

**Max deaths relative to population**
```{r max_deaths, echo = TRUE}
US_state_totals %>% slice_max(deaths_per_thou, n = 5) %>% select(deaths_per_thou, cases_per_thou, everything())
```

**Min deaths relative to population**
```{r min_deaths, echo = TRUE}
US_state_totals %>% slice_min(deaths_per_thou, n = 5) %>% select(deaths_per_thou, cases_per_thou, everything())
```

**Max cases relative to population**
```{r Max_cases, echo=TRUE}
US_state_totals %>% slice_max(cases_per_thou, n = 5) %>% select(deaths_per_thou, cases_per_thou, everything())
```

**Min cases relative to population**
```{r min_cases, echo=TRUE}
US_state_totals %>% slice_min(cases_per_thou, n = 5) %>% select(deaths_per_thou, cases_per_thou, everything())
```

Below I have also generated a bar plot indicating the states with highest incidence of Covid (cases per thousand people).  
The fill in the barplot represents deaths per thousand people for additional context.

```{r bar_plot, echo = TRUE}
# Bar plot indicating states with highest incidence of Covid (cases per thousand people).  
# The fill represents deaths per thousand for additional context.

max_cases <- US_state_totals %>% slice_max(cases_per_thou, n = 20) %>% select(deaths_per_thou, cases_per_thou, everything())

ggplot(max_cases, aes(fill = deaths_per_thou, x = reorder(Province_State, -cases_per_thou), y = cases_per_thou))+geom_bar(position="stack", stat="identity") + theme(axis.text.x = element_text(angle=90, hjust = 1))

```

### Modeling

Below is a linear model indicating the prediction for death as a function of cases in red compared with actual deaths as a function of cases in blue.
```{r modeling, echo = TRUE}
mod_death_v_case <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

US_state_with_pred <- US_state_totals %>% mutate(pred = predict(mod_death_v_case))

US_state_with_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

To capture the difference in outcomes relative to the above linear model, I have included the bar plot below to illustrate the deviation of deaths per thousand people by state and province.

The plot is ordered from left to right, best to worst outcomes respectively.

```{r modeling_delta, echo = TRUE}
US_state_death_delta <- US_state_with_pred %>% mutate(death_delta = deaths_per_thou - pred)

ggplot(US_state_death_delta, aes(x = reorder(Province_State, death_delta), y = death_delta))+geom_bar(position="stack", stat="identity") + theme(axis.text.x = element_text(angle=90, hjust = 1))
```

### Observations

From the time series data, it can be seen the number of deaths and cases was consistently lowest during summer months and highest during winter.

The highest number of cases relative to the size of the population occurred in Rhode Island, and the minimum occurred in American Samoa.  If considering only populations greater than 100,000 then Maryland would have had the lowest recorded cases per thousand people.

The highest incidence of death relative to poplulation occurred in Arizona and Oklahoma with just over 4.5 deaths per thousand people, and the lowest (if considering populations greater than 100,000) occurred in the Virgin Islands with just over 1.2 deaths per thousand people.

When analyzing the linear model and delta between predicted and actual deaths per thousand people, Alaska and Utah outcomes were ranked the best, an Oklahoma and Arizona outcomes were the worst.

From these observations, many important **follow up questions** are presented, which include but are not limited to:

* What are the contributing factors to seasonal immunosuppression that can be influenced through health and lifestyle changes?
* How did measures to prevent transmission and death differ seasonally?
* What was Alaska and Utah doing that may be transversalized and deployed in Oklahoma and Arizona to help reduce negative outcomes and excess deaths?

**Questions regarding data integrity:**

* How did standards for event labeling and data capture differ between states and municipalities?
* What incentives existed for hospitals or businesses to identify sickness or death as covid related?
* Did incentives help or hurt the reliability of data and possibly contribute to over or under reporting?

**Note regarding bias:**

The biggest factor influencing the way I approached the analysis and findings is hindsight coupled with newly published information regarding the effectiveness of measures to prevent infection, transmission and death.  

Also, hospital incentives to report new cases, as well as the lack of distinction between deaths ***with*** covid and resulting ***from*** covid were in mind during this study.  A much broader analysis, with more specific data from professionals and practices recording the most detailed information would be needed to neutralize the suspected outcomes I injected into this report. 

